---
title: "performance_of_other_algorithms"
author: "Jelena Meyer and Hansjörg Neth"
date: "2023-11-10"
output: html_document
---
```{r}
require(stats)
require(FFTrees)
```

# build regression algorithm with whole dataset

```{r}
#View(titanic)
model <- glm(survived ~., data=titanic, family="binomial")
summary(model)
```

# get accuracy statistic
```{r}
# build algorithm:
reg_log <- predict(model, titanic, type = 'response')
class_pred <- ifelse(reg_log > 0.5, 1, 0)
head(class_pred)
# get confusion matrix
confusion_matrix <- table(Predicted = class_pred, Actual = titanic$survived)
print(confusion_matrix)
# calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(accuracy)
```
# add missing data (use methods that have been built before)

```{r}
## therefore include function that adds NAs in dataset:
### Define Loop functions (for multiple simulations): ----
replace_values <- function(data, cols, amount, replacement = NA, levels_amount = NULL) {


    # Verify inputs ----
    testthat::expect_true(is.data.frame(data), info = "Data should be a dataframe.") # check that data is a data frame.
    testthat::expect_true(all(cols %in% names(data)), info = "All column names should be present in the data.") # check that columns is/are all variables in the data frame.
    testthat::expect_true(length(cols) == length(amount), info = "Number of columns and percentages have to match.") # check if columns and percentages are of the same length.
    testthat::expect_true(all(is.numeric(amount) & (amount >= 0) ), info = "All amounts should be numeric values.") # check that percentages are all numbers between 0 and 1, or higher (than treat as number of to be replaced values.
    testthat::expect_true(is.character(replacement) | is.numeric(replacement) | is.logical(replacement) | is.na(replacement), info = "Replacement value should be of a valid data type (character, nummeric, logical or NA).") # check that replacement is a valid data type.
    testthat::expect_true(is.null(levels_amount) | is.list(levels_amount), info = "levels_amount should be a list") # check that levels_amount is a list if it's not NULL.
    if (!is.null(levels_amount)) {
      testthat::expect_true(all(names(levels_amount) %in% cols), info = "All names in levels_amount have to correspond to a column in the data.") # check if all elements in levels_amount correspond to a column in the data frame.
      for (col in names(levels_amount)) {
        if (is.factor(data[[col]])) {
          testthat::expect_true(all(names(levels_amount[[col]]) %in% levels(data[[col]])), info = "All levels in levels_amount have to be present in the corresponding column of the data.") # check if all sub-elements in each element are valid levels in the corresponding factor variable.

        }

        else if (is.character(data[[col]])) {
          testthat::expect_true(all(names(levels_amount[[col]]) %in% unique(data[[col]])), info = "All levels in levels_amount have to be present in the corresponding column of the data.")

        }

      }

    }
    # main: ----

    # loop function over all columns that are inserted in form of a vector and corresponding percentages:
    for (i in seq_along(cols)) {

      col <- cols[i]
      perc <- amount[i]


      # Check if the specific column is a factor and has levels_amount defined:
      if (is.factor(data[[col]]) || is.character(data[[col]]) && !is.null(levels_amount) && col %in% names(levels_amount)) {

        # if all these conditions apply for the column, code is executed:
        # get list of different replacement percentages of levels in column:
        lev_amount <- levels_amount[[col]]

        # loop over levels in current column:
        for (cat in names(lev_amount)) {

          # get replacement percentage for current category:
          replace_perc <- lev_amount[[cat]]

          # get rows in column for current category:
          rows <- which(data[[col]] == cat)

          # Calculate how many values should be replaced in category:
          num_replace <-  ifelse(replace_perc <= 1, round(replace_perc * length(rows), 0), replace_perc)

          # Use sample to replace specified percentage of category with replacement input:
          replace_rows <- sample(rows, size = num_replace[1],  replace = FALSE)
          data[replace_rows, col] <- replacement

        }

      } else {

        # Calculate how many values should be replaced:
        num_values <- nrow(data)
        num_replace <-  ifelse(amount <= 1, round(amount * num_values, 0), amount)

        # Use sample to replace specified percentage with replacement input:
        replace_rows <- sample(1:num_values,size = num_replace[1],  replace = FALSE)
        data[replace_rows, col] <- replacement

      }

    }

    # Output: ----

    return(data) # as data frame.

  } # replace_values().
```

```{r}
titanic_20_NA <-replace_values(data = titanic, cols = c("class",    "age" ,     "sex"), amount = c(0.20, 0.20, 0.20), replacement = NA, levels_amount = NULL)
```


```{r}
# calculate log_reg with missing data

# alternative: 
# using default, which will include data anyway
# Ordinary regression with listwise exclusion of missing data
reg.fit <- glm(survived ~., data=titanic_20_NA)
summary(reg.fit)
```


```{r}
# or:
  
# build algorithm with listwise deletion (default):
reg_log_20_NA <- predict(model, titanic_20_NA, type = 'response')
class_pred_20_NA <- ifelse(reg_log_20_NA > 0.5, 1, 0)
head(class_pred_20_NA)
# get confusion matrix
confusion_matrix_20_NA <- table(Predicted = class_pred_20_NA, Actual = titanic$survived)
print(confusion_matrix_20_NA)
# calculate accuracy
accuracy_20_NA <- sum(diag(confusion_matrix_20_NA)) / sum(confusion_matrix_20_NA)
print(accuracy_20_NA) # so gehts nicht einfach, weill dann einfach nur alle NAs rausgeschmissen und nicht mehr vorhergesagt werden und die accuracy als Konsequenz tatsächlich hochgeht!

```

# somehow handle NA data in some way to be able to use log reg on the new datasets

# imputation
```{r}
require(mice)

# convert character variables into factor
titanic_20_NA <- data.frame(lapply(titanic_20_NA, function(x) if(is.character(x)) as.factor(x) else x))

```

```{r}
# having a look at imputed data
summary(titanic_20_NA)
str(titanic_20_NA) #important: factor variables defined as factors!
md.pattern(titanic_20_NA)

```

```{r}

# Multiple Imputation with the mice package
#=========================================

# 1. Imputation

imp.data <- mice (data = titanic_20_NA, m = 20, maxit = 10, seed = 123, print=FALSE)

# Which methods were used?
imp.data
#====================================================

# Checking if convergence was achieved
plot(imp.data)

# 2. Regression for each imputed data set

reg.fit.mi <- with(imp.data, glm(survived ~class + age + sex,   family = "binomial"))
reg.fit.mi


# 3. Pooling the results
pool.fit <- pool(reg.fit.mi)
summary(pool.fit)


```


# get accuracy performance!

```{r}
# Apply the logistic regression model to each imputed data set
predictions <- lapply(1:20, function(m) {
  dataset <- complete(imp.data, m)
  predict(model, new_data = dataset, type = "response")
})

# Average the predictions
avg_predictions <- Reduce("+", predictions) / 20

# Create binary classification
class_pred_imputed <- ifelse(avg_predictions > 0.5, 1, 0)

# Assuming you have the actual survival data in a vector named 'actual_survived'
confusion_matrix_imputed <- table(Predicted = class_pred_imputed, Actual = titanic$survived)
print(confusion_matrix_imputed)

# Calculate accuracy
accuracy_imputed <- sum(diag(confusion_matrix_imputed)) / sum(confusion_matrix_imputed)
print(accuracy_imputed)

```








# compare performance of best log reg NA handling method on different NA percentages with the performance of other algorithms

